---
title: "Causal Inference group file"
author: "Owen Mcwilliams"
date: "2024-03-27"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

input(What is the file for)

# Research Question

Our question (in words): This study aims at investigating the causal
effect of anti-hypertensive medication (BPMEDS) on the risk of Death

## Outcome notation

Y(1) is the outcome (death) given that the participant took
anti-hypertensive medication (BPMEDS=1) Y(0) is the outcome (death)
given that the participant did not take anti-hypertensive medication
(BPMEDS=0)

# Data

## Load Data

```{r}
fram = read.csv("framingham_assignment.csv")
```

## Data cleaning

```{r}
#creating our complete cases analysis data 
cc_fram = na.omit(fram)
sum(is.na(cc_fram))
```


## Load packages

```{r, message=F}
## Load packages
library(dagitty)
library(ggplot2)
library(GGally)
library(gridExtra)
library(dplyr)
library(ggdag)
library(cobalt)
library(survey)
library(MatchIt)
```

# Missing Data

## EDA to check which mechanism

Want to understand if the missingngess is MCAR, MAR, MNAR

```{r}
dim(fram[!complete.cases(fram),])

na_count = sum(is.na(fram$BPMEDS))+
sum(is.na(fram$EDUC))+
sum(is.na(fram$BMI))+
sum(is.na(fram$HEARTRTE)) 

na_count
na_count / dim(fram)[1]
```

There are 193 rows containing NAs (194 NAs in total). So 4.38% of our 
participants have missing data. Given that this is less than 5% of our whole 
dataset, the complete case analysis should give valid results. However we 
will explore the mechanism of missingness within BPMEDS as there are 61 missing
values, so it would be good to ensure that the missingness will not affect our 
results. 

## Missingngess of BPMEDS

```{r}
bpmeds_na_index = which(is.na(fram$BPMEDS))
bpmeds_na = fram[bpmeds_na_index,]
print("NO BPMEDS")
summary(bpmeds_na[,c(2,3,4,5,7,8,9,10)])
print("ALL DATA")
summary(fram[,c(2,3,4,5,7,8,9,10)])
```

Conclusion: There is no extreme difference in distribution for each explanatory 
variable when comparing between the full dataset and the dataset only comprising 
of participants that have NAs for BPMEDS. Therefore, our plan of action is to use
complete case analysis (remove any rows comprising of NAs). 

## BPMEDS vs SEX, AGE, SYSBP

Full dataset vs na dataset for SEX
```{r}
table(fram$SEX)
prop.table(table(fram$SEX))

table(bpmeds_na$SEX)
prop.table(table(bpmeds_na$SEX))
```


## Conclusion for analysis change

# Study protocol

1)  Eligibility criteria: Framingham community (in 1948) present in 1956
2)  Exposure definition: use of anti-hypertensive medication (BPMEDS) vs
    not using anti-hypertensive medication (BPMEDS)
3)  Assignment procedures: No randomisation - assigned based on medical
    condition?
4)  Follow-up period: 24 years Examined every 6 years
5)  Outcome definition: Binary outcome, death=1, alive=0 (at 1972)
6)  Causal contrast of interest: E(Y(1)) - E(Y(0)) (population level
    risk difference)
7)  Analysis plan: G-computation vs Propensity scores *CHOOSE WHICH ONE*

# DAG

Missing variables: EDUC, CURSMOKE, HEARTRTE, DIABP
EDA for showing why missing variables are missing

```{r}
cor(fram$SYSBP, fram$DIABP)
```

## plotting DAG

```{r, fig.width=6, fig.height=3}
dag=dagify(
  #firstly specify the relationships
  BPMEDS ~ SEX + SYSBP + AGE + BMI,
  DEATH ~ ANGINA + HOSPMI + STROKE + CVD + HYPERTEN ,
  ANGINA ~ BPMEDS,
  HOSPMI ~ BPMEDS,
  STROKE ~ BPMEDS,
  HYPERTEN ~ SYSBP,
  CVD ~ BPMEDS + BMI + AGE,
  HYPERTEN ~ BPMEDS,
  exposure = "BPMEDS",
  outcome = "DEATH",
  labels=c(
    DEATH = "DEATH",
    ANGINA = "ANGINA",
    HOSPMI = "HOSPMI",
    BPMEDS = "BPMEDS",
    EDUC = "EDUC",
    AGE = "AGE",
    HYPERTEN = "HYPERTEN",
    SYSBP = "SYSBP",
    DIABP = "DIABP",
    BPMEDS = "BPMEDS",
    CURSMOKE = "CURSMOKE",
    CVD = "CVD",
    BMI = "BMI",
    SEX = "SEX",
    STROKE = "STROKE",
    HEARTRTE = "HEARTRTE"
  )
)

# Set dag layout
theme_set(
  theme_dag() %+replace%
    # also add some additional styling
    theme(
      legend.position = "bottom",
      strip.text.x = element_text(margin = margin(2, 0, 2, 0, "mm"))
    )
)


dag_plot = ggdag(dag,use_labels = "label", text = FALSE, 
      text_size = 2, node_size = 7, layout = "sugiyama")

dag_plot

# Save as JPEG
ggsave("dag_img.jpeg", plot = dag_plot, width = 6, height = 3, dpi = 300)

```

Reword this!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!************

Mention: only focusing on the main directed arrows in our DAG as there are 
a lot of variables meaning there will be associations everywhere. These 
will overlap and make it difficult to precisely state our adjustment set. 


## checking DAG 
Here we are calculating our minimal adjustment set and ensuring that there is 
d-separation between BPMEDS and DEATH. 

```{r}
dag_tidy <- dag |>
  tidy_dagitty()

#minimal adjustment set
adjustmentSets(dag)
```

Testing the DAG to see if we missed any directed edges
```{r}
r.val <- localTests(dag, na.omit(fram))

## Adjust using Holm-Bonferroni correction
## Retain only those tests with adjusted p-value < .05
r.final = r.val[p.adjust(r.val$p.value) < 0.05, ]

r.final[order(r.final$estimate, decreasing = T), ]
```

Mention that adapting the DAG based on the above plot will not change the 
adjustment set. None are associations between BPMEDS and other covariates (only
association is between BPMEDS and DEATH which is our causal relationship of 
interest).


## final adjustment set

Final adjustment set is {AGE, BMI, SYSBP}

Mention that we don't want to adjust on the mediators (event variables) because
they show the indirect effect of BPMEDS on DEATH.

The adjustment set reaches d-separation because we have adjusted for all of 
the forks between BPEMDS and DEATH, whilst not adjusting for the chains as 
these are mediators, and there are no colliders for BPMEDS and DEATH.

# G-computation (main method)

## Fit the model

With G-computation we will start by fitting a logistic model where our outcome 
variable is the binary event of death, the explanatory variables are our 
causal variable of interest (BPMEDS) along with our minimal adjustment set 
decided from our DAG. 


Will decide the best model using information criterion

```{r}
model1_gc = glm(DEATH ~ BPMEDS + BMI + SYSBP + AGE,
              family = binomial,
              data = cc_fram)

summary(model1_gc)

```


```{r}
cc_fram_1 = cc_fram
cc_fram_1$BPMEDS = 1

E_Yhat_1 = predict(model1_gc, newdata = cc_fram_1, type = "response")
EY1 = mean(E_Yhat_1)

cc_fram_0 = cc_fram
cc_fram_0$BPMEDS = 0

E_Yhat_0 = predict(model1_gc, newdata = cc_fram_0, type = "response")
EY0 = mean(E_Yhat_0)

paste("E(Y(1)) =", round(EY1,3), ", E(Y(0)) =", round(EY0,3))
paste("Average Treatment Effect (ATE) =", round(EY1-EY0,3))

```

## Check assumptions

(exchangeability and positivity) 

```{r, fig.width=4, fig.height=3}
theme_set(theme_grey())

par(mfrow = c(1, 3))

p_age = ggplot(data=cc_fram,
       aes(x=as.factor(BPMEDS), y=AGE, fill = as.factor(BPMEDS)) ) + 
  geom_boxplot() +
  labs(x = "BPMEDS") +
  theme(legend.position = "none")

p_sys = ggplot(data=cc_fram,
       aes(x=as.factor(BPMEDS), y=SYSBP, fill = as.factor(BPMEDS)) ) + 
  geom_boxplot() +
  labs(x = "BPMEDS") +
  theme(legend.position = "none")

p_bmi = ggplot(data=cc_fram,
       aes(x=as.factor(BPMEDS), y=BMI, fill = as.factor(BPMEDS))) + 
  geom_boxplot( ) +
  labs(x = "BPMEDS") +
  theme(legend.position = "none")

p_age
p_sys
p_bmi

ggsave("age_boxplot.jpeg", plot = p_age, width = 4, height = 3, dpi = 300)
ggsave("sys_boxplot.jpeg", plot = p_sys, width = 4, height = 3, dpi = 300)
ggsave("bmi_boxplot.jpeg", plot = p_bmi, width = 4, height = 3, dpi = 300)


```

!!!!!!!!!!!!!!!!!!!!!!!!!!! 
WORK OUT HOW TO CHECK THE CONDITIONAL EXCHANGEABILITY ASSUMPTION

# Propensity scores (sensitivity analysis)

## Fit propensity scores

```{r}
ps1.model = glm(BPMEDS ~ BMI + SYSBP + AGE, family=binomial, data=cc_fram)
summary(ps1.model)
```

```{r}
cc_fram$ps1 = fitted.values(ps1.model)
summary(cc_fram$ps1)
```

## Fit weights model 

```{r}
cc_fram$ipw1 <- (cc_fram$BPMEDS==1)/cc_fram$ps1+
                (cc_fram$BPMEDS==0)/(1-cc_fram$ps1)
summary(cc_fram$ipw1)
```


### Check assumptions
First checking exchangeability 

```{r}
vars1 <- c("BMI" , "AGE" , "SYSBP", "SEX", "DIABP", "CURSMOKE", "EDUC",
           "HEARTRTE")
covariates <-cc_fram[, vars1]  
bt = bal.tab(covariates, treat = cc_fram$BPMEDS, weights = cc_fram$ipw1, 
        method = "weighting", un=TRUE)
l_plot = love.plot(covariates, treat = cc_fram$BPMEDS, weights = cc_fram$ipw1,
          method = "weighting",   binary = "std", threshold = .1)

l_plot

ggsave("love_plot.jpeg", plot = l_plot, width = 6, height = 3, dpi = 300)
```
After fitting our propensity score model there remain imbalances across the 
confounders, indicating that the conditional exchangeability assumption is 
violated. 

Now exploring the positivity assumption:

```{r}
positivity_plot = ggplot(data=cc_fram) + 
  geom_density(alpha=0.2, aes(x=ps1, 
 fill=as.factor(BPMEDS))) + xlim(c(0, 1)) +
  labs(x = "Propensity score", y = "Density") +
    guides(
    fill = guide_legend(order = 1, title = "BPMEDS exposure")
  ) +
  theme(legend.position = c(0.85, 0.85)) 

positivity_plot

ggsave("positivit_plot.jpeg", plot = positivity_plot, width = 6, height = 3, dpi = 300)
```

There is a sufficient overlap so the positivity assumption holds. 

### ATE 

```{r}
# weighted mean outcome in treatment
mean1.ipw <-weighted.mean(cc_fram$DEATH[cc_fram$BPMEDS==1], 
                          w=cc_fram$ipw1[cc_fram$BPMEDS==1])
mean1.ipw

# weighted mean outcome in untreated
mean0.ipw <-weighted.mean(cc_fram$DEATH[cc_fram$BPMEDS==0], 
                          w=cc_fram$ipw1[cc_fram$BPMEDS==0])
mean0.ipw

# ATE estimate
mean1.ipw - mean0.ipw
```

```{r}
# Calculate ATE with robust Standard errors
d.w <- svydesign(~1, weights = cc_fram$ipw1, data = cc_fram)
fit.ipw <- svyglm(DEATH ~ BPMEDS, design = d.w)
summary(fit.ipw)
```
```{r}
confint(fit.ipw)
```


## Fit
matching model explore different matching ideas 

```{r}
m = matchit(BPMEDS ~ BMI + SYSBP + AGE, family=binomial, data=cc_fram)
summary(m)
```

```{r}
md = match.data(m)
table(md$BPMEDS, md$DEATH)
prop.table(table(md$BPMEDS, md$DEATH), margin=1)
```
### ATT
```{r}
att = prop.table(table(md$BPMEDS, md$DEATH), margin=1)[2,2] - 
      prop.table(table(md$BPMEDS, md$DEATH), margin=1)[1,2]
att
```



# Appendix

## EDA

Talk about imbalance between BPMEDS proportions for whole dataset

### DAG

### Propensity scores
